---
title: "Getting started with matrispace"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with matrispace}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

# Overview

`matrispace` is the R package companion to MatriSpace. It mirrors the same
workflow structure used in the web app: data input, matrisome profiling, and
feature analysis.

The package is designed for local, reproducible analysis of spatial
transcriptomics data with a focus on extracellular matrix biology.

```{r load-package}
library(matrispace)
library(Seurat)
library(ggplot2)
```

# Load data

You can start from a Seurat object or convert a SpatialExperiment object with
`convert_spe_to_seurat()`.

```{r load-data, eval=FALSE}
seurat_obj <- readRDS("path/to/your/spatial_object.rds")
```

For examples and testing, the package also ships with a small Seurat object:

```{r load-test-data, eval=FALSE}
test_path <- system.file("testdata", "test_seurat.rds", package = "matrispace")
seurat_obj <- readRDS(test_path)
```

# Prepare the object

`prepare_object()` runs the standard MatriSpace preprocessing steps:

- gene/signature translation
- matrisome scoring
- ECM niche classification
- ECM niche scoring

```{r preprocess, eval=FALSE}
seurat_obj <- prepare_object(seurat_obj, verbose = TRUE)
```

# Matrisome profiling

The main profiling functions mirror the MatriSpace app workflow.

```{r profiling, eval=FALSE}
seurat_obj <- score_matrisome(seurat_obj)
seurat_obj <- annotate_ecm_niches(seurat_obj)
seurat_obj <- score_matrisome_pair_coexpression(seurat_obj)
```

Built-in reference objects are available directly:

```{r reference-data}
data(matrisome_db)
data(matrisome_signatures)
data(ecm_signatures)
data(matrisome_pairs)

nrow(matrisome_db)
length(matrisome_signatures)
names(ecm_signatures)
```

# Visualization

You can generate the same style of ECM-focused spatial views used throughout
MatriSpace.

```{r plots, eval=FALSE}
plot_spatial_feature(seurat_obj, "Collagens")
plot_spatial_blend(seurat_obj, c("COL1A1", "FN1"))
```

# Differential expression

The package now exposes two complementary differential-expression workflows.

## Within-sample spatial correction

Use `find_spatially_corrected_ecm_markers()` when comparing niches or
compartments within a tissue section and spatial autocorrelation should be
modeled explicitly.

```{r spatial-de, eval=FALSE}
spatial_de <- find_spatially_corrected_ecm_markers(
  seurat_obj,
  group.by = "CompositionCluster_CC",
  ident.1 = "CC1",
  ident.2 = "CC3"
)
```

## Cross-sample pseudobulk testing

Use `find_ecm_de_samples()` with `pseudobulk = TRUE` when comparing conditions
across biological samples.

```{r pseudobulk-de, eval=FALSE}
pseudobulk_de <- find_ecm_de_samples(
  seurat_obj,
  condition.col = "condition",
  condition.1 = "Inflamed",
  condition.2 = "Non-inflamed",
  pseudobulk = TRUE,
  sample.col = "sample_id",
  cluster.col = "CompositionCluster_CC",
  blocking.col = "patient_id"
)
```

# Spatial statistics

`matrispace` also provides direct access to local and global spatial analyses.

```{r spatial-stats, eval=FALSE}
lisa_df <- compute_lisa(seurat_obj, "COL1A1", "gene", "FN1", "gene")
morans <- compute_morans_i(seurat_obj)
```

# Related resources

- [MatriSpace web app](https://izzilab-matrispace.share.connect.posit.cloud/)
- [MatriSpace app repository](https://github.com/Theayomideo/matrispace-app)
- [The Matrisome Project](https://sites.google.com/uic.edu/matrisome/home)
