---
title: "MatriSpace: Spatially-Resolved ECM Gene Expression Patterns"
author: "Oshinjo A, Chen D, Petrov P, Izzi V, Naba A"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  warning = FALSE,
  message = FALSE,
  cache = TRUE
)
```

```{r load-packages, include=FALSE}
library(matrispace)
library(Seurat)
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyr)
```

```{r load-data, include=FALSE}
# Load test Seurat object
test_paths <- c(
  system.file("testdata", "test_seurat.rds", package = "matrispace"),
  "../inst/testdata/test_seurat.rds",
  "inst/testdata/test_seurat.rds"
)
test_path <- NULL
for (p in test_paths) {
  if (file.exists(p) && p != "") {
    test_path <- p
    break
  }
}
seurat_obj <- readRDS(test_path)
seurat_obj <- UpdateSeuratObject(seurat_obj)
seurat_obj <- prepare_object(seurat_obj, verbose = FALSE)
```

# Overview

**matrispace** analyzes the extracellular matrix (matrisome) in spatial transcriptomics data. This demo shows:

- ECM niche classification
- Matrisome expression mapping
- Spatial co-localization (LISA)
- Spatial co-expression analysis of matrisome pairs

**Test dataset:** Spatial transcriptomics sample with matrisome annotation

---

# ECM Niche Annotation

Each spot is classified into an ECM niche (e.g. Interstitial, Basement
membrane) using ScType.

```{r ecm-distribution}
table(seurat_obj$ecm_niche)
```

```{r ecm-spatial, fig.height=5}
ecm_colors <- palette_ecm(unique(seurat_obj$ecm_niche))
coords <- GetTissueCoordinates(seurat_obj)
plot_df <- data.frame(
  imagecol = coords$imagecol,
  imagerow = coords$imagerow,
  niche = seurat_obj$ecm_niche
)

ggplot(plot_df, aes(x = imagecol, y = imagerow, color = niche)) +
  geom_point(size = 1.3) +
  scale_color_manual(values = ecm_colors, name = "ECM niche") +
  scale_y_reverse() +
  coord_fixed() +
  theme_void() +
  labs(title = "ECM niche classification") +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "right")
```

---

# Matrisome Expression

UCell scores for matrisome categories visualized spatially.

```{r matrisome-spatial, fig.height=10}
p1 <- plot_spatial_feature(seurat_obj, "Collagens", pt.size = 1.3)
p2 <- plot_spatial_feature(seurat_obj, "ECM Glycoproteins", pt.size = 1.3)
p3 <- plot_spatial_feature(seurat_obj, "ECM Regulators", pt.size = 1.3)
p4 <- plot_spatial_feature(seurat_obj, "ECM-affiliated Proteins", pt.size = 1.3)

(p1 | p2) / (p3 | p4)
```

## Dual Normalization View

Robust-normalized and log-scaled scores side-by-side:

```{r matrisome-dual, fig.height=5}
seurat_obj$collagens_robust_score <- seurat_obj@meta.data[["Collagens"]]
seurat_obj$collagens_log_scaled_score <- log1p(seurat_obj@meta.data[["Collagens"]])

plot_matrisome(seurat_obj, "Collagens", "collagens", pt.size = 1.3)
```

---

# Spatial Co-localization (LISA)

LISA (Local Indicators of Spatial Association) identifies where two features cluster together spatially.

```{r lisa-compute}
# Analyze COL4A1 vs COL1A1 spatial relationship
lisa_results <- compute_lisa(
  seurat_obj,
  feature1 = "COL4A1", feature1_type = "gene",
  feature2 = "COL1A1", feature2_type = "gene"
)
lisa_results$id <- seurat_obj$ecm_niche

table(lisa_results$LISA)
```

```{r lisa-plot, fig.width=12, fig.height=6}
plot_lisa(
  lisa_results,
  ann_col = "id",
  f1_name = "COL4A1",
  f2_name = "COL1A1"
)
```

**Interpretation:**

| Pattern | Meaning |
|---------|---------|
| High-High (red) | COL4A1 high, surrounded by COL1A1 high |
| High-Low (orange) | COL4A1 high, surrounded by COL1A1 low |
| Low-High (purple) | COL4A1 low, surrounded by COL1A1 high |
| Low-Low (blue) | Both low |

---

# Two-Feature Blend

Visualize co-expression with a 2D color gradient:

```{r blend-plot, fig.width=12, fig.height=4}
plot_spatial_blend(
  seurat_obj,
  features = c("COL4A1", "COL1A1"),
  feature1_name = "COL4A1",
  feature2_name = "COL1A1"
)
```

---

# Matrisome Pair Co-expression Analysis

Matrisome-pair scoring measures spatial co-expression across adjacent spots for unique heterotypic MatriComDB pairs. It does not infer signaling direction or communication activity.

```{r lr-score, cache=TRUE}
seurat_obj <- score_lr_activity(seurat_obj, verbose = FALSE)
cat("Matrisome pairs scored:", nrow(seurat_obj[["LRACTIVITY"]]))
```

## Enrichment by ECM Niche

```{r lr-enrichment}
lr_stats <- find_lr_enrichment(seurat_obj, group_by = "ecm_niche")
```

### Top Interactions per Niche

```{r lr-top}
get_top <- function(cluster, n = 5) {
  lr_stats %>%
    filter(cluster == !!cluster) %>%
    slice_max(avg_log2FC, n = n) %>%
    select(feature, avg_log2FC, perc_difference)
}
```

**Interstitial ECM niche:**
```{r}
get_top("Interstitial_ECM")
```

**Basement membrane ECM niche:**
```{r}
get_top("Basement_ECM")
```

## Heatmap

```{r lr-heatmap, fig.height=7}
top_lr <- lr_stats %>%
  group_by(cluster) %>%
  slice_max(avg_log2FC, n = 8) %>%
  ungroup()

ggplot(top_lr, aes(x = cluster, y = reorder(feature, avg_log2FC), fill = avg_log2FC)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b", midpoint = 0) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8),
    panel.grid = element_blank()
  ) +
  labs(x = NULL, y = NULL, fill = "log2FC",
       title = "Top Enriched Matrisome Pairs")
```

---

# Differential ECM Expression

Identify ECM genes differentially expressed between ECM niches.

## Markers for one niche

Compare Interstitial ECM vs all other niches:

```{r de-single}
interstitial_markers <- find_ecm_markers(
  seurat_obj,
  ident.1 = "Interstitial_ECM",
  group.by = "ecm_niche",
  verbose = FALSE
)

head(interstitial_markers[, c("gene", "avg_log2FC", "p_val_adj", "subcategory")])
```

## All niche markers

Find ECM markers for each niche simultaneously:

```{r de-all}
all_markers <- find_all_ecm_markers(
  seurat_obj,
  group.by = "ecm_niche",
  verbose = FALSE
)

# Top 3 markers per niche
all_markers %>%
  group_by(cluster) %>%
  slice_max(avg_log2FC, n = 3) %>%
  select(cluster, gene, avg_log2FC, subcategory)
```

## Category-Specific Analysis

Filter to specific matrisome categories (e.g., collagens only):

```{r de-category}
collagen_markers <- find_ecm_markers(
  seurat_obj,
  ident.1 = "Basement_ECM",
  group.by = "ecm_niche",
  categories = "Collagens",
  verbose = FALSE
)
collagen_markers[, c("gene", "avg_log2FC", "p_val_adj")]
```

## Visualization

```{r de-heatmap, fig.height=6}
top_de <- all_markers %>%
  filter(p_val_adj < 0.05) %>%
  group_by(cluster) %>%
  slice_max(avg_log2FC, n = 6) %>%
  ungroup()

ggplot(top_de, aes(x = cluster, y = reorder(gene, avg_log2FC), fill = avg_log2FC)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b", midpoint = 0) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  ) +
  labs(x = NULL, y = NULL, fill = "log2FC", title = "Top ECM markers by niche")
```

---

# Summary

| Function | What it does |
|----------|--------------|
| `annotate_ecm_niches()` | Classify spots into ECM niches |
| `score_matrisome()` | Calculate matrisome category scores |
| `plot_spatial_feature()` | Map any feature spatially |
| `compute_lisa()` | Spatial co-localization statistics |
| `plot_lisa()` | Visualize LISA clusters |
| `plot_spatial_blend()` | Two-feature overlay |
| `score_lr_activity()` | Matrisome-pair spatial co-expression scoring |
| `find_lr_enrichment()` | Matrisome-pair enrichment per group |
| `find_ecm_markers()` | DE between two groups |
| `find_all_ecm_markers()` | DE for all clusters |

---

# Session Info

```{r session-info}
si <- sessionInfo()
cat("R version:", si$R.version$version.string, "\n")
cat("Platform:", si$platform, "\n\n")
cat("Key packages:\n")
pkgs <- c("matrispace", "Seurat", "SeuratObject", "ggplot2", "UCell")
for (pkg in pkgs) {
  if (pkg %in% names(si$otherPkgs)) {
    cat(" ", pkg, ":", as.character(packageVersion(pkg)), "\n")
  }
}
```
